---
title: "Case-control specialized functions in genmixr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case-control specialized functions in genmixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(genmixr)
```

# Draft note

Draft vignette. Equation/page references will be finalized after textbook verification.

TODO: Add exact textbook equation numbers and pages after verification.

# Overview

The specialized case-control functions are focused convenience functions for users
who want one calculation at a time. The full functions,
`cc_power()` and `cc_mssn()`, are still the
main interface for combined workflows.

This vignette groups the specialized functions by calculation type.

```{r}
g_case <- c(0.25, 0.50, 0.25)
g_ctrl <- c(0.36, 0.48, 0.16)
```

# Genotype chi-square functions

The genotype-only misclassification functions use model-based conditional
genotype frequencies and return genotype chi-square MSSN or power.

This evaluated example verifies that the vignette can run:

```{r}
cc_chisq_power_genotype_misclassification_1p(
  N_case = 200,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  e = 0.02,
  verbose = FALSE
)$power
```

```{r, eval = FALSE}
cc_chisq_mssn_genotype_misclassification_1p(
  power = 0.80,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  e = 0.02,
  verbose = FALSE
)
```

# Trend test functions

Trend functions use genotype scores `w`, with the default scores
`c(0, 1, 2)`.

```{r, eval = FALSE}
cc_trend_power_locus_heterogeneity(
  N_case = 200,
  alpha = 0.05,
  g_case_assoc = g_case,
  g_ctrl = g_ctrl,
  pi = 0.80,
  w = c(0, 1, 2),
  verbose = FALSE
)

cc_trend_mssn_locus_heterogeneity(
  power = 0.80,
  alpha = 0.05,
  g_case_assoc = g_case,
  g_ctrl = g_ctrl,
  pi = 0.80,
  w = c(0, 1, 2),
  verbose = FALSE
)
```

# Genotype misclassification functions

Use the 2p or 3p functions when the genotype misclassification model is known
in advance and only genotype chi-square results are needed.

```{r, eval = FALSE}
cc_chisq_power_genotype_misclassification_2p(
  N_case = 200,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  e1 = 0.01,
  e2 = 0.02,
  verbose = FALSE
)

cc_chisq_power_genotype_misclassification_3p(
  N_case = 200,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  e01 = 0.02,
  e02 = 0.01,
  e03 = 0.005,
  verbose = FALSE
)
```

# Locus heterogeneity functions

These functions apply the case-frequency adjustment
`pi * g_case_assoc + (1 - pi) * g_ctrl` before computing the requested test.

```{r, eval = FALSE}
cc_chisq_power_locus_heterogeneity(
  N_case = 200,
  alpha = 0.05,
  g_case_assoc = g_case,
  g_ctrl = g_ctrl,
  pi = 0.80,
  verbose = FALSE
)

cc_chisq_mssn_locus_heterogeneity(
  power = 0.80,
  alpha = 0.05,
  g_case_assoc = g_case,
  g_ctrl = g_ctrl,
  pi = 0.80,
  verbose = FALSE
)
```

# Differential 3p functions

Differential 3p functions use separate case and control misclassification
matrices.

```{r, eval = FALSE}
cc_chisq_power_genotype_misclassification_differential_3p(
  N_case = 200,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  case_e01 = 0.02,
  case_e02 = 0.01,
  case_e03 = 0.005,
  ctrl_e01 = 0.01,
  ctrl_e02 = 0.005,
  ctrl_e03 = 0.002,
  verbose = FALSE
)

cc_chisq_mssn_genotype_misclassification_differential_3p(
  power = 0.80,
  alpha = 0.05,
  prev = 0.10,
  pd = 0.30,
  R2 = 1.80,
  MOI = "D",
  case_e01 = 0.02,
  case_e02 = 0.01,
  case_e03 = 0.005,
  ctrl_e01 = 0.01,
  ctrl_e02 = 0.005,
  ctrl_e03 = 0.002,
  verbose = FALSE
)
```
