---
title: "Case-control full functions in genmixr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case-control full functions in genmixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(genmixr)
```

# Methodological map

Model-based penetrances follow Chapter 1 Section 1.4.2, Eqs. 1.6--1.7
(p. 13) of Gordon, Finch, and Kim (2020). The genotype chi-square and trend
tests use Eqs. 1.22 and 1.24 (pp. 26--27), with the trend statistic defined in
Eqs. 1.20--1.21 (p. 24). Locus heterogeneity follows Eq. 2.16 (p. 88), and
the trend-test formulation under heterogeneity is in Eqs. 5.29a--b
(pp. 287--288). Genotype-error matrices are described by Eqs. 2.5--2.7
(pp. 57--58).

# Overview

The full case-control interface has two main functions:

- `cc_power()` computes power for fixed case/control sample sizes.
- `cc_mssn()` computes minimum sample size necessary (MSSN) for a target power.

Both functions follow the same workflow:

1. Construct baseline case and control genotype frequencies.
2. Optionally apply locus heterogeneity.
3. Optionally apply genotype misclassification.
4. Compute genotype chi-square and genotype trend calculations.

The examples below use a small model-free setup:

```{r}
g_case <- c(0.25, 0.50, 0.25)
g_ctrl <- c(0.36, 0.48, 0.16)
```

# Model-free, no locus heterogeneity, no misclassification

This evaluated example verifies that the vignette runs.

```{r}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  geno_misclass = "none",
  verbose = FALSE
)$tests$genotypes$power
```

# Model-free with locus heterogeneity

```{r, eval = FALSE}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  locus_het = TRUE,
  pi = 0.80,
  verbose = FALSE
)
```

# Model-free with locus heterogeneity and 3p misclassification

```{r, eval = FALSE}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  locus_het = TRUE,
  pi = 0.80,
  geno_misclass = "3p",
  e01 = 0.02,
  e02 = 0.01,
  e03 = 0.005,
  verbose = FALSE
)
```

# Model-based with locus heterogeneity and 3p misclassification

```{r, eval = FALSE}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_based",
  prev = 0.05,
  pd = 0.30,
  R2 = 1.80,
  MOI = "M",
  locus_het = TRUE,
  pi = 0.80,
  geno_misclass = "3p",
  e01 = 0.02,
  e02 = 0.01,
  e03 = 0.005,
  verbose = FALSE
)
```

# MSSN example

```{r, eval = FALSE}
cc_mssn(
  power = 0.80,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  locus_het = TRUE,
  pi = 0.80,
  geno_misclass = "3p",
  e01 = 0.02,
  e02 = 0.01,
  e03 = 0.005,
  verbose = FALSE
)
```

# Differential 3p, explicit parameters

```{r, eval = FALSE}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  geno_misclass = "diff3p",
  diff_source = "explicit",
  case_e01 = 0.02,
  case_e02 = 0.01,
  case_e03 = 0.005,
  ctrl_e01 = 0.01,
  ctrl_e02 = 0.005,
  ctrl_e03 = 0.002,
  verbose = FALSE
)
```

# Differential 3p, multiplier from case parameters

```{r, eval = FALSE}
cc_power(
  N_case = 200,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_ctrl,
  geno_misclass = "diff3p",
  diff_source = "case",
  diff_multiplier = 0.50,
  case_e01 = 0.02,
  case_e02 = 0.01,
  case_e03 = 0.005,
  verbose = FALSE
)
```
