---
title: "Getting started with paweh"
author: "paweh package authors"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Getting started with paweh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(paweh)
```

# What problem does `paweh` solve?

`paweh` supports prospective power and minimum-sample-size-necessary (MSSN)
calculations for genetic association study designs. It is intended for the
stage at which a researcher is deciding what data to collect, how many
subjects or trios are needed, and how sensitive that design may be to model
assumptions.

The relevant design inputs differ across studies. They can include sample
size, allele frequency, genotype relative risk, genotype or phenotype
misclassification, locus heterogeneity, family-based transmission,
extreme-phenotype sampling, and correlation among quantitative traits.
These quantities are assumptions or design targets, not estimates produced
from raw genomic data by `paweh`.

# Power and MSSN answer inverse questions

- **Power:** given a sample size and model, what probability of rejection is
  expected under the specified alternative?
- **MSSN:** given a target power and model, how many cases, controls, trios, or
  total individuals are required?

For example, suppose pilot work supplies conditional genotype probabilities
for cases and controls. A compact power calculation is:

```{r tiny-example}
g_case <- c(0.25, 0.50, 0.25)
g_control <- c(0.36, 0.48, 0.16)

starter_power <- cc_power(
  N_case = 500,
  alpha = 0.05,
  input_mode = "model_free",
  g1 = g_case,
  g0 = g_control,
  k = 1,
  verbose = FALSE
)

starter_power$tests$genotypes$power
```

The corresponding `cc_mssn()` calculation reverses the design question by
replacing `N_case` with a target `power`.

# Three study-design families

## Case-control association

Cases and controls are compared using a genotype chi-square or trend test.
The canonical `cc_power()` and `cc_mssn()` interfaces support model-based and
model-free inputs, along with compatible heterogeneity and error modifiers.

## Affected-child trios

The transmission disequilibrium test (TDT) compares transmitted and
non-transmitted alleles in affected-child trios. `tdt_power()` and
`tdt_mssn()` report power or required trio counts under distinct no-error,
phenotype-misclassification, and locus-heterogeneity scenarios.

## Quantitative traits

For continuous traits, `paweh` supports one-way genotype-group ANOVA,
threshold-selected genotype chi-square designs, and multivariate designs
using Pillai MANOVA or joint threshold selection.

# Model-based and model-free inputs

These modes locate the scientific assumptions in different places.

For **case-control** studies, model-based input uses prevalence, disease-allele
frequency, genotype relative risk, and mode of inheritance. Model-free input
uses supplied conditional genotype probabilities: `g1` for cases or affected
subjects and `g0` for controls or unaffected subjects.

For the **TDT**, model-based input uses `pd`, `prev`, `R1`, `R2`, and
`delta_prime`. Model-free input starts from expected transmission and
non-transmission counts (`ET` and `ENT`), with the associated number of trios
needed for MSSN calculations.

Model-free input is useful when these quantities come from pilot data, prior
publications, external calculations, or direct design assumptions. It does
not remove assumptions; it moves them into the supplied conditional
quantities.

# Where should I start?

| Scientific design | Canonical starting functions | Detailed vignette |
|---|---|---|
| Case-control association | `cc_power()`, `cc_mssn()` | Case-Control Study Design |
| Affected-child trios | `tdt_power()`, `tdt_mssn()` | TDT Study Design |
| Continuous single-trait QTL | `qtl_anova_power()`, `qtl_anova_mssn()` | Quantitative-Trait Study Design |
| Extreme-trait QTL | `qtl_threshold_chisq_power()`, `qtl_threshold_chisq_mssn()` | Quantitative-Trait Study Design |
| Multivariate quantitative traits | `qtl_multivariate_power_full()`, `qtl_multivariate_mssn_full()` | Quantitative-Trait Study Design |

# Why Bioconductor?

`paweh` is being prepared for Bioconductor because it addresses prospective
design questions in genetic association research. These calculations can
complement downstream genomic analysis by examining operating characteristics
before data generation or collection. The package does not currently claim
sequencing-file, genomic-range, or `SummarizedExperiment` integration.

# Published validation philosophy

The remaining vignettes distinguish three uses of literature:

- **Published-example reproduction:** a reported analytic result is evaluated
  with a matching `paweh` design.
- **Published methodological context:** a paper explains an important risk or
  method, but its exact analysis is not claimed to be reproduced.
- **Motivating published application:** an application motivates a design
  question while using a different statistical procedure.

The study-design vignettes reproduce validated examples across case-control,
TDT, and multivariate quantitative-trait settings and then extend those fixed
designs with sensitivity calculations.

# Installation

Once `paweh` is available through Bioconductor, the standard installation path
will be:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("paweh")
```

No installation is performed while this vignette is built.

# Session information

```{r session-info}
sessionInfo()
```
