---
title: "Sequencing-Based Genetic Study Design with PAWEH"
author: "paweh package authors"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{6. Sequencing-Based Genetic Study Design with PAWEH}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(paweh)
```

# Overview

`paweh` supports prospective analytic power and minimum-sample-size-necessary
(MSSN) calculations for two scientifically different sequencing designs. The
inputs describe a planned experiment; these functions do not process FASTQ,
BAM, or VCF files and do not perform downstream association testing.

For case-control sequencing designs, the pathway is:

> true genotype -> sequencing read model -> deterministic genotype call ->
> 3 by 3 transition matrix -> Ahn/Chapman--Nam trend NCP -> power or MSSN.

For affected-child sequencing trios, the pathway is:

> latent Mendelian trio state -> raw read-count likelihood -> efficient
> information -> Kim TDT1-NGS NCP -> power or MSSN.

The case-control calculation therefore propagates uncertainty in called
genotypes. TDT1-NGS instead retains raw-read likelihood information and does
not hard-call the three genotypes. These are not interchangeable methods.

# Two sequencing workflows in `paweh`

| Feature | Case-control NGS | TDT1-NGS |
|---|---|---|
| Sampling unit | Cases and controls | Complete father-mother-affected-child trios |
| Sequencing information | Called-genotype probabilities | Raw read-count likelihood |
| Association kernel | Ahn/Chapman--Nam trend NCP | Kim efficient-information NCP |
| Public power function | `cc_ngs_power()` | `tdt_ngs_power()` |
| Public MSSN function | `cc_ngs_mssn()` | `tdt_ngs_mssn()` |
| Locus heterogeneity | Available | Not currently implemented |

Both designs currently use equal fixed coverage and a symmetric public
per-read error parameter. Neither uses simulation.

# Exact published TDT1-NGS replication: the Kim AIS example

Kim (2015) used adolescent idiopathic scoliosis (AIS) to illustrate a
low-coverage trio design. The motivating variant, rs1400180, had a reported
global minor-allele frequency near 0.349; the design calculation used 0.35.
The published assumptions were HWE, symmetric read error 0.025, genome-wide
`alpha = 5e-8`, a multiplicative effect reported as OR at least 2, and target
power 0.90. In the TDT1-NGS parameterization this effect is `R1 = 2`, with
`R2 = R1^2`.

```{r kim-ais-replication}
kim_4x <- tdt_ngs_mssn(
  power = 0.90,
  pd = 0.35,
  R1 = 2,
  coverage = 4,
  seq_error = 0.025,
  alpha = 5e-8,
  verbose = FALSE
)

kim_25x <- tdt_ngs_mssn(
  power = 0.90,
  pd = 0.35,
  R1 = 2,
  coverage = 25,
  seq_error = 0.025,
  alpha = 5e-8,
  verbose = FALSE
)

data.frame(
  coverage = c("4x", "25x"),
  published_trios = c(654, 416),
  paweh_trios = c(kim_4x$MSSN_trios, kim_25x$MSSN_trios)
)
```

Under Kim's published HWE design with MAF 0.35, symmetric read error 0.025,
genome-wide alpha $5\times10^{-8}$, and multiplicative effect `R1 = 2`,
PAWEH exactly reproduces the reported integer requirements of 654 complete
trios at 4x and 416 at 25x for 90% power. This statement applies to this
specific published design and is not a claim about every model in the paper.

Kim's full Table 3 pairs three target powers with three effect sizes. All six
integer results reproduce exactly:

```{r kim-table-3}
kim_table3 <- data.frame(
  coverage = rep(c(4, 25), times = 3),
  power = rep(c(0.90, 0.70, 0.10), each = 2),
  R1 = rep(c(2.0, 1.8, 1.5), each = 2),
  published_trios = c(654, 416, 716, 456, 733, 466)
)

kim_table3$paweh_trios <- c(
  kim_4x$MSSN_trios,
  kim_25x$MSSN_trios,
  vapply(3:6, function(i) {
    tdt_ngs_mssn(
      power = kim_table3$power[i], pd = 0.35, R1 = kim_table3$R1[i],
      coverage = kim_table3$coverage[i], seq_error = 0.025,
      alpha = 5e-8, verbose = FALSE
    )$MSSN_trios
  }, numeric(1))
)
kim_table3
```

# TDT1-NGS power and MSSN

For Kim's one-degree-of-freedom multiplicative model, the NCP is

\[
\lambda=N\{\log(R_1)\}^2 I_{eff},
\]

where $N$ is the number of complete trios and $I_{eff}$ is the
nuisance-adjusted per-trio information obtained from the raw-read likelihood
under the null. The public interface does not require users to manipulate the
full information matrix.

```{r tdt-ngs-workflow}
tdt_seq_power <- tdt_ngs_power(
  N = 5000,
  pd = 0.325,
  R1 = 1.2,
  coverage = 12,
  seq_error = 0.005,
  alpha = 5e-8,
  verbose = FALSE
)

tdt_seq_mssn <- tdt_ngs_mssn(
  power = 0.80,
  pd = 0.325,
  R1 = 1.2,
  coverage = 12,
  seq_error = 0.005,
  alpha = 5e-8,
  verbose = FALSE
)

data.frame(
  quantity = c("Power at 5,000 trios", "Required trios", "Total individuals"),
  value = c(
    tdt_seq_power$power,
    tdt_seq_mssn$MSSN_trios,
    tdt_seq_mssn$total_individuals
  )
)
```

The MSSN calculation first finds the target one-degree-of-freedom NCP and then
solves analytically for $N$. It is not an iterative simulation over possible
trio counts.

# TDT sequencing sensitivity

A small coverage-by-error grid helps identify whether a proposed design is in
a steep low-depth region or near a higher-depth plateau.

```{r tdt-ngs-power-plot, fig.cap="TDT1-NGS power across equal fixed coverage for two symmetric per-read error assumptions."}
plot_ngs_power(
  design = "tdt",
  coverage = c(2, 4, 8, 12, 20),
  seq_error = c(0.005, 0.01),
  target_power = 0.80,
  N = 5000, pd = 0.325, R1 = 1.2, alpha = 5e-8
)
```

```{r tdt-ngs-mssn-plot, fig.cap="Required complete trios across equal fixed coverage for two symmetric per-read error assumptions."}
plot_ngs_mssn(
  design = "tdt",
  coverage = c(2, 4, 8, 12, 20),
  seq_error = c(0.005, 0.01),
  power = 0.80, pd = 0.325, R1 = 1.2, alpha = 5e-8
)
```

Low-to-moderate depth gains can be large, whereas gains at higher depth may
plateau. The effect of sequencing error remains setting-dependent and should
not be assumed negligible. The current fixed-depth TDT1-NGS implementation
does not support a design in which every member of every trio is modeled at
exactly 1x coverage under the full nuisance parameterization. This is a
specific identifiability boundary of the implemented model, not a claim that
1x sequencing is universally unusable.

# Published case-control framework: the Ahn example

Ahn et al. (2007) studied ordinary genotype misclassification, not sequencing
depth. The paper supplies genotype-error probabilities directly. PAWEH
implements Ahn Equation (1), its analytic MSSN rearrangement, and the published
trend scores T001 = `(0,0,1)`, T011 = `(0,1,1)`, and T012 = `(0,1,2)`.

The stronger numerical example uses prevalence 0.005, marker and disease
allele frequencies 0.30, $D'=0.5$, and relative risks 2 and 3. The following
conditional frequencies were independently reconstructed from Ahn's genetic
model during package validation. Adjacent-genotype errors of 5% correspond to
the PAWEH three-parameter matrix with `e01 = e02 = 0.05` and `e03 = 0`.

```{r ahn-published-method-replication}
ahn_case <- c(0.398125, 0.472500, 0.129375)
ahn_control <- c(0.490461683417085, 0.419736180904523, 0.089802135678392)

ahn_no_error <- cc_power(
  N_case = 1000, alpha = 0.001,
  input_mode = "model_free",
  g1 = ahn_case, g0 = ahn_control,
  k = 1, w = c(0, 1, 2),
  geno_misclass = "none", verbose = FALSE
)

ahn_adjacent_error <- cc_power(
  N_case = 1000, alpha = 0.001,
  input_mode = "model_free",
  g1 = ahn_case, g0 = ahn_control,
  k = 1, w = c(0, 1, 2),
  geno_misclass = "3p", e01 = 0.05, e02 = 0.05, e03 = 0,
  verbose = FALSE
)

data.frame(
  scenario = c("No error", "No error", "5% adjacent error", "5% adjacent error"),
  test = c("T012", "Genotype chi-square", "T012", "Genotype chi-square"),
  published_approximate_power = c(0.88, 0.81, 0.79, 0.70),
  paweh_power = c(
    ahn_no_error$tests$trend$power,
    ahn_no_error$tests$genotypes$power,
    ahn_adjacent_error$tests$trend$power,
    ahn_adjacent_error$tests$genotypes$power
  )
)
```

This is a **Published-method replication**, not a CC-NGS sequencing-depth
replication. PAWEH implements Ahn's Equation (1) exactly and closely reproduces
the paper's ordinary genotype-misclassification power example.

# The PAWEH CC-NGS sequencing bridge

Ahn takes $P(\text{observed genotype}\mid\text{true genotype})$ as an input.
PAWEH adds an upstream analytic bridge from sequencing assumptions to those
probabilities. For true genotype $g\in\{0,1,2\}$, alternate-read count $X$,
and fixed depth $v$,

\[
X\mid G=g,V=v\sim\operatorname{Binomial}(v,q_g),\qquad
(q_0,q_1,q_2)=(\epsilon,0.5,1-\epsilon).
\]

A deterministic maximum-likelihood rule calls genotype $k$. Each transition
probability

\[
E_{gk}(v,\epsilon)=P(\text{called genotype}=k\mid\text{true genotype}=g)
\]

is obtained by finite summation over all possible read counts. Rows of $E$
are true genotypes and columns are called genotypes. This coverage-to-matrix
bridge is a PAWEH composition; it was not proposed in the Ahn paper.

```{r cc-transition-matrix}
ngs_genotype_error_matrix(coverage = 4, seq_error = 0.005)
```

# CC-NGS power and MSSN

The public case-control interfaces construct true case and control genotype
probabilities from the disease model, apply the sequencing transition matrix,
and then use the Ahn/Chapman--Nam trend NCP.

```{r cc-ngs-workflow}
cc_seq_power <- cc_ngs_power(
  N_case = 1000,
  alpha = 5e-8,
  prev = 0.05,
  pd = 0.30,
  R2 = 1.8,
  coverage = 4,
  seq_error = 0.005,
  MOI = "M",
  k = 1,
  verbose = FALSE
)

cc_seq_mssn <- cc_ngs_mssn(
  power = 0.80,
  alpha = 5e-8,
  prev = 0.05,
  pd = 0.30,
  R2 = 1.8,
  coverage = 4,
  seq_error = 0.005,
  MOI = "M",
  k = 1,
  verbose = FALSE
)

data.frame(
  quantity = c(
    "Power at 1,000 cases", "Required cases", "Required controls", "Total MSSN"
  ),
  value = c(
    cc_seq_power$power,
    cc_seq_mssn$MSSN_case,
    cc_seq_mssn$MSSN_ctrl,
    cc_seq_mssn$MSSN_total
  )
)
```

Sequencing uncertainty is upstream of the published trend NCP. The MSSN
calculation remains analytic after inversion of the target NCP.

# Why zero per-read error is not perfect at finite depth

Setting `seq_error = 0` eliminates incorrect base reads, but it does not make
finite-depth heterozygote calls perfect. For a true heterozygote,

\[
X\sim\operatorname{Binomial}(v,0.5).
\]

At low $v$, every observed read can come from the same allele. Such an
observation can be more likely under a homozygous genotype, creating call
uncertainty even with no per-read sequencing error. This probability decreases
with depth but is nonzero at every finite depth.

```{r zero-error-matrices}
list(
  coverage_2 = ngs_genotype_error_matrix(2, 0),
  coverage_10 = ngs_genotype_error_matrix(10, 0)
)
```

# CC sequencing sensitivity

```{r cc-ngs-power-plot, fig.cap="Case-control sequencing trend power across equal fixed coverage for two symmetric per-read error assumptions."}
plot_ngs_power(
  design = "cc",
  coverage = c(2, 4, 8, 12, 20),
  seq_error = c(0.005, 0.01),
  target_power = 0.80,
  N_case = 1000, alpha = 5e-8,
  prev = 0.05, pd = 0.30, R2 = 1.8, MOI = "M", k = 1
)
```

```{r cc-ngs-mssn-plot, fig.cap="Required case sample size across equal fixed coverage for two symmetric per-read error assumptions."}
plot_ngs_mssn(
  design = "cc",
  coverage = c(2, 4, 8, 12, 20),
  seq_error = c(0.005, 0.01),
  power = 0.80, alpha = 5e-8,
  prev = 0.05, pd = 0.30, R2 = 1.8, MOI = "M", k = 1
)
```

These are deterministic scenario curves, not uncertainty bands. A robust
design should be examined over scientifically plausible depth and error
values rather than only at a single optimistic point.

# CC locus heterogeneity

Case-control sequencing designs can dilute the associated-case distribution
with the control distribution:

\[
g_{case,H}=\pi g_{case}+(1-\pi)g_{control}.
\]

Here `pi = 1` retains the full locus-associated case distribution and `pi = 0`
removes the locus-specific contrast. The master switch is strict: when
`locus_het = FALSE`, `pi` must remain 1.

```{r cc-ngs-heterogeneity}
cc_heterogeneous <- cc_ngs_power(
  N_case = 1000, alpha = 0.05,
  prev = 0.05, pd = 0.30, R2 = 1.8,
  coverage = 4, seq_error = 0.005,
  MOI = "M", k = 1,
  locus_het = TRUE, pi = 0.75,
  verbose = FALSE
)

data.frame(
  pi = cc_heterogeneous$locus_het$pi,
  lambda = cc_heterogeneous$lambda,
  power = cc_heterogeneous$power
)
```

# The exact `pi = 0` boundary

At `locus_het = TRUE, pi = 0`, case and control genotype distributions are
identical. Therefore the NCP is zero and power equals `alpha`. If requested
power exceeds `alpha`, no finite sample size exists.

The plotting-data interface preserves this scientific boundary instead of
inventing a very large finite MSSN:

```{r cc-ngs-null-boundary}
pi_boundary <- plot_ngs_mssn(
  design = "cc",
  coverage = 4,
  seq_error = 0.005,
  return_data = TRUE,
  power = 0.80, alpha = 0.05,
  prev = 0.05, pd = 0.30, R2 = 1.8, MOI = "M", k = 1,
  locus_het = TRUE, pi = c(1, 0.75, 0.50, 0)
)

pi_boundary[, c("pi", "MSSN_case", "finite_mssn", "status")]
```

The `pi = 0` row remains visible with `MSSN_case = NA`,
`finite_mssn = FALSE`, and status `"no finite MSSN"`.

# Fixed-depth assumptions and limitations

Current sequencing APIs assume one equal fixed coverage value for the relevant
study members. They do not yet integrate over a realistic empirical or
budget-generated coverage distribution in which a locus may be 0x in some
samples, 1x or 2x in others, and 4x or greater elsewhere. Variable/BGE
coverage-distribution modeling is deferred.

| Assumption | CC-NGS | TDT1-NGS |
|---|---|---|
| Locus | Single biallelic | Single biallelic |
| Sampling | Cases and controls | Complete affected-child trios |
| Genetic model | M, dominant, or recessive trend scores | Multiplicative |
| Parental HWE/random mating | Not applicable | Assumed by current public design |
| Coverage | Equal fixed depth | Equal fixed depth |
| Public read error | Symmetric | Symmetric |
| Sequencing analysis | Deterministic genotype calls | Raw-read latent-state likelihood |
| Case/control observation error | Nondifferential | Not applicable |
| Locus heterogeneity | Optional canonical mixture | Not implemented |
| Simulation | None | None |

Deferred capabilities include variable/BGE coverage, simulation validation,
cost-depth optimization, TDT sequencing heterogeneity, and a sequencing Shiny
module. The existing dashboard covers the non-sequencing study-design modules.

# A practical design workflow

1. Choose an unrelated case-control or affected-child trio design.
2. Specify allele-frequency and effect assumptions from defensible prior data.
3. Choose candidate fixed coverage and per-read error scenarios.
4. Calculate power at the available sample size.
5. Calculate MSSN for the target power.
6. Plot coverage and sequencing-error sensitivity.
7. For case-control designs, stress-test plausible locus heterogeneity.
8. Report the model, fixed-depth assumption, sampling unit, error assumption,
   and sensitivity range with the design recommendation.

# References

Ahn K, Haynes C, Kim W, St Fleur R, Gordon D, Finch SJ. The effects of SNP
genotyping errors on the power of the Cochran-Armitage linear trend test for
case/control association studies. *Annals of Human Genetics*.
2007;71(2):249--261.
<https://doi.org/10.1111/j.1469-1809.2006.00318.x>.

Chapman DG, Nam JM. Asymptotic power of chi square tests for linear trends in
proportions. *Biometrics*. 1968;24(2):315--327.
<https://doi.org/10.2307/2528035>.

Gordon D, Finch SJ, Kim W. *Heterogeneity in Statistical Genetics: How to
Assess, Address, and Account for Mixtures in Association Studies*. Springer;
2020. <https://doi.org/10.1007/978-3-030-61121-7>.

Kim W. Transmission disequilibrium tests based on read counts for low-coverage
next-generation sequence data. *Human Heredity*. 2015;80(1):36--49.
<https://doi.org/10.1159/000434645>.

# Session information

```{r session-info}
sessionInfo()
```
