---
title: "LRTae power and sample-size design"
author: "paweh package authors"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{7. LRTae Power and Sample-Size Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(paweh)
```

# What is LRTae?

LRTae is a likelihood-ratio framework for case-control genetic association
studies in which phenotype classification, genotype classification, or both
may be measured with error. A subset of participants can receive more accurate
or gold-standard measurements. These validation data inform the error process
and contribute information about the association.

The basic problem is easy to state:

> true phenotype and genotype -> fallible measured phenotype and genotype ->
> potentially distorted apparent association

Validation provides a second path:

> gold-standard measurements on a subset -> information about
> misclassification -> improved efficient information for the association
> test

`paweh` implements the **prospective design** calculations `lrtae_ncp()`,
`lrtae_power()`, and `lrtae_mssn()`. They calculate expected operating
characteristics before data collection. The package does not implement the
observed-data iterative likelihood fitting or association test from the
historical LRTae spreadsheet and software.

# Four validation tiers

Each planned participant belongs to one mutually exclusive measurement tier.

| Tier | Phenotype | Genotype | Fixed-design argument | MSSN allocation argument |
|---|---|---|---|---|
| Fallible only | Fallible | Fallible | `n_fallible` | `prop_fallible` |
| Phenotype validated | Gold standard | Fallible | `n_pheno_validated` | `prop_pheno_validated` |
| Genotype validated | Fallible | Gold standard | `n_geno_validated` | `prop_geno_validated` |
| Both validated | Gold standard | Gold standard | `n_both_validated` | `prop_both_validated` |

The four fixed counts sum to the total sample size. The four MSSN proportions
must be nonnegative and sum to one.

> **Matrix orientation is important:** for both misclassification matrices,
> rows are **true** categories and columns are **observed** categories.

# A reproducible prospective design

Consider a research team planning a case-control association study. Routine
phenotype ascertainment is imperfect, routine genotyping has a small
classification error, and gold-standard clinical adjudication and confirmatory
genotyping are too expensive for every participant.

The following deterministic inputs come from a historical LRTae development
workbook. They are used here as a reproducible worked design, not as observed
patient data and not as a realized NCP benchmark.

```{r historical-inputs}
p_control <- c(0.5645454545, 0.3739393939, 0.0615151515)
p_case <- c(0.36, 0.48, 0.16)
q <- c(0.5, 0.5)

phenotype_error <- matrix(
  c(0.98, 0.02,
    0.04, 0.96),
  nrow = 2,
  byrow = TRUE
)
genotype_error <- matrix(
  c(0.965, 0.025, 0.010,
    0.025, 0.950, 0.025,
    0.010, 0.025, 0.965),
  nrow = 3,
  byrow = TRUE
)

model_inputs <- list(
  true_control_genotype_frequencies = p_control,
  true_case_genotype_frequencies = p_case,
  true_sampling_proportions = q,
  phenotype_misclassification_matrix = phenotype_error,
  genotype_misclassification_matrix = genotype_error
)
tier_counts <- list(
  n_fallible = 248,
  n_pheno_validated = 135,
  n_geno_validated = 72,
  n_both_validated = 45
)
```

The sampling proportions are in control/case order. The genotype-frequency
vectors use the same three-category order as the rows and columns of
`genotype_error`.

# Computing the NCP

```{r historical-ncp}
historical_ncp <- do.call(lrtae_ncp, c(model_inputs, tier_counts))
historical_ncp
```

The noncentrality parameter (NCP) summarizes the expected separation between
the alternative and null distributions. Larger values imply greater expected
ability to detect association, holding the reference distribution fixed.

Technically, the implementation evaluates

\[
\lambda = \Delta^\mathsf{T} I_{\mathrm{eff}} \Delta,
\]

where \(\Delta\) contrasts the free control and case genotype frequencies and
\(I_{\mathrm{eff}}\) is their efficient information after nuisance parameters
are eliminated by a Schur complement. This is an expected-information design
quantity; it is not the likelihood-ratio statistic from one simulated or
observed data set.

# Computing power

```{r historical-power}
historical_power <- do.call(
  lrtae_power,
  c(model_inputs, tier_counts, list(alpha = 0.05))
)
historical_power
```

For three genotype categories, the test has `3 - 1 = 2` degrees of freedom.
The returned object records power, NCP, degrees of freedom, significance level,
critical value, all four tier counts, and total sample size.

# Computing minimum sample size

The workbook allocation corresponds to proportions 248/500, 135/500, 72/500,
and 45/500. The same planned allocation can be used to ask how many total
participants are required for 80% power.

```{r historical-mssn}
historical_mssn <- do.call(
  lrtae_mssn,
  c(
    model_inputs,
    list(
      target_power = 0.80,
      alpha = 0.05,
      prop_fallible = 248 / 500,
      prop_pheno_validated = 135 / 500,
      prop_geno_validated = 72 / 500,
      prop_both_validated = 45 / 500
    )
  )
)
historical_mssn
```

`MSSN_total` is the minimum integer total under this allocation rule. The
Hamilton largest-remainder rule assigns integer tier counts that sum exactly
to the selected total, so the achieved proportions can differ slightly from
the requested proportions. The returned achieved power is therefore normally
just above, rather than exactly equal to, the target.

# Why validation allocation matters

Here four designs have the same total sample size (`N = 220`) and identical
biological and error assumptions. Only the allocation across validation tiers
changes. The low-validation design retains a small number in every validation
tier; complete absence of validation can leave the error model unidentified
and its information matrix singular.

```{r validation-comparison}
allocations <- data.frame(
  design = c(
    "Low validation", "Phenotype focused", "Genotype focused", "Balanced"
  ),
  n_fallible = c(198, 154, 154, 110),
  n_pheno_validated = c(9, 44, 11, 33),
  n_geno_validated = c(9, 11, 44, 33),
  n_both_validated = c(4, 11, 11, 44),
  stringsAsFactors = FALSE
)

design_results <- lapply(seq_len(nrow(allocations)), function(index) {
  counts <- as.list(allocations[index, 2:5])
  do.call(lrtae_power, c(model_inputs, counts, list(alpha = 0.05)))
})
allocations$N_total <- rowSums(allocations[2:5])
allocations$NCP <- vapply(design_results, function(x) x$ncp, numeric(1))
allocations$Power <- vapply(design_results, function(x) x$power, numeric(1))
allocations
```

```{r validation-figure, fig.cap="Prospective LRTae power for four validation allocations with the same total sample size and error assumptions."}
barplot(
  allocations$Power,
  names.arg = allocations$design,
  las = 2,
  ylim = c(0, 1),
  ylab = "Prospective power",
  col = "#3B7EA1"
)
abline(h = 0.80, lty = 2, col = "#A33A2B")
```

Validation is therefore not merely a data-cleaning exercise: it contributes
statistical information. The comparison does not establish that one allocation
is universally optimal. Phenotype and genotype validation can trade off
differently under other frequencies, error matrices, costs, or effect sizes.

# Reading the misclassification matrices

For the phenotype matrix,

\[
\begin{pmatrix}
0.98 & 0.02 \\
0.04 & 0.96
\end{pmatrix},
\]

98% of true controls are observed as controls and 2% as cases; 4% of true
cases are observed as controls and 96% as cases. Again, **true categories are
rows and observed categories are columns**.

The genotype matrix is read the same way. For example, a participant in the
first true genotype category is observed in categories one, two, and three
with probabilities 0.965, 0.025, and 0.010. Each row must sum to one, and its
category order must match both genotype-frequency vectors.

# When to use these functions

The current LRTae functions are appropriate when:

- a case-control genetic association study is being planned;
- expected true control and case genotype distributions are available;
- phenotype or genotype measurement may be imperfect;
- validation subsamples are planned; and
- the investigator wants an NCP, power, or minimum total sample size.

Important limitations are:

- these are prospective calculations, not observed-data association fitting;
- error matrices are treated as known design assumptions;
- validity depends on the assumed error model;
- validation allocation affects both information and identifiability;
- this implementation is not LTTae; and
- it is not a sequencing/read-level error model.

# References

Barral S, Haynes C, Stone M, Gordon D. LRTae: improving statistical power for
genetic association with case/control data when phenotype and/or genotype
misclassification errors are present. *BMC Genetics*. 2006;7:24.
<https://doi.org/10.1186/1471-2156-7-24>.

Ji F, Yang Y, Haynes C, Finch SJ, Gordon D. Computing asymptotic power and
sample size for case-control genetic association studies in the presence of
phenotype and/or genotype misclassification errors. *Statistical Applications
in Genetics and Molecular Biology*. 2005;4(1):Article 37.
<https://doi.org/10.2202/1544-6115.1184>.

Gordon D, Finch SJ, Kim W. *Heterogeneity in Statistical Genetics: How to
Assess, Address, and Account for Mixtures in Association Studies*. Springer;
2020. <https://doi.org/10.1007/978-3-030-61121-7>.

# Session information

```{r session-info}
sessionInfo()
```
