---
title: "Transmission disequilibrium test study design"
author: "paweh package authors"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Transmission disequilibrium test study design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(paweh)
```

# Why use an affected-child trio design?

The transmission disequilibrium test (TDT) compares an allele transmitted to
an affected child with the alternative parental allele that was not
transmitted. This within-family comparison can avoid confounding from
population stratification that affects unrelated case-control comparisons.

Let `ET` and `ENT` denote expected transmitted and non-transmitted counts. The
one-degree-of-freedom TDT non-centrality parameter is based on their contrast,

\[
\lambda = \frac{(ET-ENT)^2}{ET+ENT}.
\]

In this vignette, `N` and every reported MSSN are numbers of affected-child
trios, not numbers of individual people.

# Power and MSSN for trios

The canonical model-based interfaces answer inverse design questions under a
common genetic model.

```{r basic-tdt}
basic_tdt_power <- tdt_power(
  N = 600, input_mode = "model_based",
  pd = 0.30, prev = 0.05, R1 = 1.5, R2 = 2.25,
  alpha = 0.05, delta_prime = 1,
  misclass_rate = 0.01, heter_rate = 0.10,
  verbose = FALSE
)

basic_tdt_mssn <- tdt_mssn(
  target_power = 0.80, input_mode = "model_based",
  pd = 0.30, prev = 0.05, R1 = 1.5, R2 = 2.25,
  alpha = 0.05, delta_prime = 1,
  misclass_rate = 0.01, heter_rate = 0.10,
  verbose = FALSE
)

data.frame(
  scenario = c("No error", "Phenotype misclassification", "Locus heterogeneity"),
  power_at_600_trios = unlist(basic_tdt_power$power),
  required_trios = unlist(basic_tdt_mssn$N),
  row.names = NULL
)
```

The canonical backend reports separate modifier scenarios. It does not claim
that the misclassification and heterogeneity components are a combined-error
analysis.

# Model-based and model-free TDT inputs

Model-based input uses `pd`, `prev`, `R1`, `R2`, and `delta_prime`. The
lower-level probability and count functions make the connection to `ET` and
`ENT` visible without requiring a long derivation.

```{r expected-counts}
gT <- tdt_expected_transmission_probability(
  pd = 0.30, prev = 0.05, R1 = 1.5, R2 = 2.25,
  delta_prime = 1, verbose = FALSE
)
gNT <- tdt_expected_nontransmission_probability(
  pd = 0.30, prev = 0.05, R1 = 1.5, R2 = 2.25,
  delta_prime = 1, verbose = FALSE
)
expected_counts <- tdt_expected_transmission_counts(
  N_star = 500,
  pd = 0.30, prev = 0.05, R1 = 1.5, R2 = 2.25,
  delta_prime = 1, pi = 1, verbose = FALSE
)

data.frame(
  quantity = c("Transmission probability", "Non-transmission probability", "ET", "ENT"),
  value = c(gT$gT_star, gNT$gNT_star, expected_counts$ET_star, expected_counts$ENT_star)
)
```

If a pilot study or external calculation already supplies expected counts,
they can be used directly:

```{r model-free-power}
count_power <- tdt_power_from_expected_counts(
  ET = expected_counts$ET_star,
  ENT = expected_counts$ENT_star,
  alpha = 0.05
)

canonical_model_free <- tdt_power(
  N = 500, input_mode = "model_free",
  ET = expected_counts$ET_star,
  ENT = expected_counts$ENT_star,
  misclass_rate = 0, heter_rate = 0,
  verbose = FALSE
)

c(lower_level_power = count_power$power,
  canonical_power = canonical_model_free$power$no_error)
```

For model-free MSSN, `n_trios` records the trio count represented by the
supplied `ET` and `ENT`. Applying heterogeneity additionally requires `pd`;
applying phenotype misclassification requires `pd` and `prev`. Model-free
inputs therefore relocate assumptions rather than eliminate them.

# Published-example reproduction: Buyske et al. (2009)

Buyske et al. examined phenotype misclassification in affected-child trios.
The validated multiplicative example uses target power 0.80, `alpha = 1e-5`,
`pd = 0.50`, prevalence 0.01, `R2 = 2.5`, `R1 = sqrt(2.5)`, maximum positive
LD, and phenotype misclassification rate 0.05.

```{r buyske-reproduction}
buyske <- tdt_mssn(
  target_power = 0.80, alpha = 1e-5,
  pd = 0.50, prev = 0.01,
  R1 = sqrt(2.5), R2 = 2.5,
  delta_prime = 1,
  misclass_rate = 0.05, heter_rate = 0,
  verbose = FALSE
)

buyske_ratio <- buyske$N$misclassification / buyske$N$no_error
data.frame(
  quantity = c("No-error MSSN", "Misclassification MSSN", "Final sample-size ratio"),
  published = c("approximately 546", "approximately 21,400", "approximately 39-fold"),
  paweh = c(buyske$N$no_error, buyske$N$misclassification, buyske_ratio)
)
```

This is an excellent **Published-example reproduction**: approximately 546
versus 21,401 affected-child trios, a 39.23-fold final sample-size ratio.
A 39.23-fold ratio is a 3,822.7% increase, not a 39.23% increase. The
`percent_increase` component stores this conventional percentage as
`100 * ((new / baseline) - 1)`; the validated final-to-baseline ratio remains
39.23-fold.

## Extending Buyske: a misclassification curve

```{r buyske-curve, fig.cap="Required affected-child trios as phenotype misclassification increases under the Buyske model."}
plot_tdt_mssn(
  x_var = "misclass_rate",
  x_values = seq(0, 0.05, by = 0.005),
  scenario = "misclassification",
  input_mode = "model_based",
  target_power = 0.80, alpha = 1e-5,
  pd = 0.50, prev = 0.01,
  R1 = sqrt(2.5), R2 = 2.5,
  delta_prime = 1, heter_rate = 0,
  title = "Required trios vs phenotype misclassification"
)
```

The steep curve identifies a fragile design region. It does not represent a
confidence interval or a probability distribution over the error rate.

## Extending Buyske: prevalence by misclassification

The optional interactive surface evaluates the same Buyske parameters over a
modest prevalence-by-error grid. It is guarded so the vignette remains
executable when Plotly is unavailable.

```{r buyske-surface, eval=requireNamespace("plotly", quietly=TRUE), fig.cap="MSSN sensitivity to prevalence and phenotype misclassification under the Buyske model."}
plot_tdt_surface3d(
  metric = "mssn", scenario = "misclassification",
  x = "prev", y = "misclass_rate",
  x_values = c(0.005, 0.01, 0.02, 0.04),
  y_values = seq(0, 0.05, length.out = 5),
  target_power = 0.80, alpha = 1e-5,
  pd = 0.50, R1 = sqrt(2.5), R2 = 2.5,
  delta_prime = 1, heter_rate = 0
)
```

This surface extends discrete calculations into a sensitivity analysis; it is
not an uncertainty distribution.

# Published-example reproduction: Chen et al. (2009)

Chen et al. studied TDT design under locus heterogeneity. Here `heter_rate` is
the heterogeneous or unlinked trio fraction, equal to `1 - pi`.

```{r chen-reproduction}
chen_no_heterogeneity <- tdt_mssn(
  target_power = 0.80, alpha = 1e-5,
  pd = 0.25, prev = 0.10,
  R1 = sqrt(1.5), R2 = 1.5,
  delta_prime = 1,
  misclass_rate = 0, heter_rate = 0,
  verbose = FALSE
)

chen_half_heterogeneous <- tdt_mssn(
  target_power = 0.80, alpha = 1e-5,
  pd = 0.25, prev = 0.10,
  R1 = sqrt(1.5), R2 = 1.5,
  delta_prime = 1,
  misclass_rate = 0, heter_rate = 0.50,
  verbose = FALSE
)

data.frame(
  heterogeneous_fraction = c(0, 0.50),
  published_required_trios = c("approximately 3430", "approximately 13721"),
  paweh_required_trios = c(
    chen_no_heterogeneity$N$heterogeneity,
    chen_half_heterogeneous$N$heterogeneity
  )
)
```

This is a near-exact **Published-example reproduction**. Small differences
reflect the paper's reported integer values and rounding.

## Extending Chen: dilution by heterogeneous trios

```{r chen-curve, fig.cap="Required affected-child trios as the heterogeneous or unlinked fraction increases under the Chen model."}
plot_tdt_mssn(
  x_var = "heter_rate",
  x_values = seq(0, 0.50, by = 0.05),
  scenario = "heterogeneity",
  input_mode = "model_based",
  target_power = 0.80, alpha = 1e-5,
  pd = 0.25, prev = 0.10,
  R1 = sqrt(1.5), R2 = 1.5,
  delta_prime = 1, misclass_rate = 0,
  title = "Required trios vs locus heterogeneity"
)
```

Heterogeneous trios dilute locus-specific signal. A flatter curve would
indicate a more robust design; a steep curve signals that modest departures
from homogeneity require substantially more families.

# Assumptions and limitations

- The design is an affected-child trio TDT using an asymptotic chi-square
  approximation.
- In the current parameterization, `delta_prime` lies in `[0, 1]` and scales
  maximum **positive** LD. Negative LD requires another allele-frequency-
  dependent normalization and is not implemented.
- `misclass_rate` and `heter_rate` describe distinct reported scenarios.
- Model-free `ET` and `ENT` are treated as supplied expected counts.
- MSSN is measured in trios.
- Curves and surfaces vary fixed assumptions; they are not uncertainty bands.

# References

Buyske S, Yang G, Matise TC, Gordon D. When a case is not a case: effects of
phenotype misclassification on power and sample size requirements for the
transmission disequilibrium test with affected child trios. *Human Heredity*.
2009;67(4):287--292. doi:10.1159/000194981.

Chen C, Yang G, Buyske S, Matise T, Finch SJ, Gordon D. Transmission
disequilibrium test power and sample size in the presence of locus
heterogeneity. *Statistical Applications in Genetics and Molecular Biology*.
2009;8:Article 44. doi:10.2202/1544-6115.1501.

Gordon D, Finch SJ, Kim W. *Heterogeneity in Statistical Genetics: How to
Assess, Address, and Account for Mixtures in Association Studies*. Springer;
2020. doi:10.1007/978-3-030-61121-7.

# Session information

```{r session-info}
sessionInfo()
```
