---
title: "Transmission Disequilibrium Test (TDT) power and sample size in genmixr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TDT power and sample size in genmixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(genmixr)
library(ggplot2)
```

# 1. Introduction

Family-based association tests, particularly the *Transmission Disequilibrium Test* (TDT), are widely used in human genetics to detect linkage and association while avoiding confounding due to population stratification.

The **genmixr** package implements an extended TDT framework that accounts for

* Phenotype misclassification
* Locus heterogeneity
* Linkage disequilibrium scaling
* Power calculation
* Sample size determination

This vignette focuses exclusively on the two main user-facing functions:

* `tdt_power_full()`
* `tdt_required_trios_full()`

These functions provide a complete workflow for designing family-based studies.

# 2. Required Libraries

```{r}
library(genmixr)
library(ggplot2)
```

# 3. Power Calculations with `tdt_power_full()`

Suppose we are planning a TDT study with 600 affected trios, a high-risk allele frequency of 0.30, and genotype relative risks 1.5 and 2.25 for the heterozygote and risk homozygote, respectively. We also assume a disease prevalence of 5%, LD scale parameter `delta_prime` = 1, and modest misclassification and heterogeneity

```{r}

tdt_power_full(
  N = 600,
  pd = 0.30,
  prev = 0.05,
  R1 = 1.5,
  R2 = 2.25,
  misclass_rate = 0.01,
  heter_rate = 0.10,
  delta_prime = 1,
  alpha = 0.05,
  verbose = TRUE
)

```

The returned object is a list with components such as:

* `no_error`: power under an ideal model with no misclassification or heterogeneity.
* `misclassification`: power under phenotype misclassification only
* `heterogeneity`: power under locus heterogeneity only

The `power_loss` element gives the absolute loss in power (relative to the no-error design) caused by misclassification and heterogeneity.

# 4. Required Sample Size with `tdt_required_trios_full()`

Now suppose instead we want to determine how many affected trios are required
to achieve 80% power at the same parameter values.

```{r}

res_trios <- tdt_required_trios_full(
  target_power = 0.80,
  pd = 0.30,
  prev = 0.05,
  R1 = 1.5,
  R2 = 2.25,
  misclass_rate = 0.01,
  heter_rate = 0.10,
  delta_prime = 1,
  alpha = 0.05,
  verbose = TRUE
)

```

The main quantities of interest are:

```{r}

res_trios$N
res_trios$percent_increase

```

where: 

* `N$no_error` is the required sample size under an ideal model.
* `N$misclassification` and `N$heterogeneity` are the required trios if the design accounts for misclassification or heterogeneity.
* `percent_increase` gives the proportional increase in sample size relative
to the no-error design (e.g. 0.42 corresponds to a 42% increase).


# 5. Real-world examples

In this section we illustrate how `tdt_power_full()` and `tdt_required_trios_full()` can be used to interpret and design family-based association studies. These examples show how one might *use* `genmixr` for real design scenarios rather than reproducing original papers exactly.

We rely on two studies:

- **Ambrosio et al. (2004)** — DRD2/DRD3 TDT in schizophrenia trios (negative results).  
   
- **Spielman et al. (1993)** — the original TDT paper on the insulin VNTR and IDDM (strong positive results).  
  


## 5.1 DRD2/DRD3 and schizophrenia in Portuguese trios

Ambrosio et al. analyzed **90 affected trios** to test DRD2/DRD3 variants in schizophrenia. They reported **no evidence of association**.

From their TDT table for the DRD2 TaqIA polymorphism:

- **Transmitted:** 72  
- **Not transmitted:** 69  

This near 50/50 split suggests low power to detect modest effects.


### 5.1.1 Approximate power under modest genetic effects

Assume the DRD2 risk allele truly had a modest effect:

- Prevalence: 1%  
- Allele frequency (`pd`): 0.30  
- Genotype relative risks: `R1 = 1.3`, `R2 = 1.3^2`

```{r}
ambrosio_power <- tdt_power_full(
  N            = 90,
  pd           = 0.30,
  prev         = 0.01,
  R1           = 1.3,
  R2           = 1.3^2,
  misclass_rate = 0,
  heter_rate    = 0,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

ambrosio_power$no_error
ambrosio_power$power_loss
```

This typically yields very low power, consistent with the negative results in Ambrosio et al.


### 5.1.2 Impact of misclassification and heterogeneity

Psychiatric phenotypes often include diagnostic uncertainty. We examine sensitivity:

```{r}
ambrosio_power_error <- tdt_power_full(
  N             = 90,
  pd            = 0.30,
  prev          = 0.01,
  R1            = 1.3,
  R2            = 1.3^2,
  misclass_rate = 0.05,
  heter_rate    = 0.20,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

ambrosio_power_error$misclassification
ambrosio_power_error$heterogeneity
ambrosio_power_error$power_loss
```

Even small misclassification fractions can markedly reduce power, especially in small studies.


### 5.1.3 Required trios for adequate power

We now ask how many trios would be required to reach **80% power** under the same assumptions:

```{r}
ambrosio_trios <- tdt_required_trios_full(
  target_power  = 0.80,
  pd            = 0.30,
  prev          = 0.01,
  R1            = 1.3,
  R2            = 1.3^2,
  misclass_rate = 0.05,
  heter_rate    = 0.20,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

ambrosio_trios$N
ambrosio_trios$percent_increase
```

These results illustrate how misclassification and heterogeneity inflate the required sample size.


## 5.2 Insulin VNTR and IDDM: the original TDT paper

Spielman, McGinnis, and Ewens introduced the TDT using the insulin gene 5′ flanking polymorphism (5′FP) in IDDM.

From their Table 5:

- **Transmitted (class 1):** 78  
- **Not transmitted:** 46  

This gives a strong χ² (~8.26, p ≈ 0.004).


### 5.2.1 Approximate effective sample size

We approximate:

- Effective trios: `N ≈ 120`  
- Allele frequency: `pd = 0.75`  
- Prevalence: 0.004  
- Genotype RR: `R1 = 1.8`, `R2 = 1.8^2`

```{r}
spielman_power <- tdt_power_full(
  N             = 120,
  pd            = 0.75,
  prev          = 0.004,
  R1            = 1.8,
  R2            = 1.8^2,
  misclass_rate = 0,
  heter_rate    = 0,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

spielman_power$no_error
```

This yields high power, consistent with their strong published result.


### 5.2.2 Robustness to misclassification and heterogeneity

```{r}
spielman_power_error <- tdt_power_full(
  N             = 120,
  pd            = 0.75,
  prev          = 0.004,
  R1            = 1.8,
  R2            = 1.8^2,
  misclass_rate = 0.02,
  heter_rate    = 0.10,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

spielman_power_error$no_error
spielman_power_error$misclassification
spielman_power_error$heterogeneity
spielman_power_error$power_loss
```

Because the true effect is strong, the TDT remains robust even with realistic error.


### 5.2.3 Designing a follow-up study

Suppose we want **90% power** to detect a smaller effect (`R1 = 1.6`):

```{r}
spielman_trios_followup <- tdt_required_trios_full(
  target_power  = 0.90,
  pd            = 0.75,
  prev          = 0.004,
  R1            = 1.6,
  R2            = 1.6^2,
  misclass_rate = 0.02,
  heter_rate    = 0.10,
  delta_prime   = 1,
  alpha         = 0.05,
  verbose       = TRUE
)

spielman_trios_followup$N
spielman_trios_followup$percent_increase
```

This demonstrates how required trios increase when targeting more modest effects.


Together, these examples show how `genmixr` can be used both to reinterpret past TDT studies and to design new trio studies under realistic assumptions.


# 6. Visualizing Sensitivity to Error Processes

To summarize the influence of phenotype misclassification and locus heterogeneity on TDT performance, **genmixr** provides four plotting utilities:

- `tdt_plot_power_misclassification()`: power vs. misclassification rate (holding heterogeneity fixed)
- `tdt_plot_power_heterogeneity()`: power vs. heterogeneity rate (holding misclassification fixed)
- `tdt_plot_sample_size_misclassification()`: required trios vs. misclassification (holding heterogeneity fixed)
- `tdt_plot_sample_size_heterogeneity()`: required trios vs. heterogeneity (holding misclassification fixed)

These plots provide a concise graphical summary of the same error processes examined numerically above.

## 6.1 Power sensitivity: misclassification

```{r}
tdt_plot_power_misclassification(
  N    = 600,
  pd   = 0.30,
  prev = 0.05,
  R1   = 1.5,
  R2   = 2.25,
  alpha = 0.05,
  delta_prime = 1,
  misclass_seq = seq(0, 0.15, by = 0.01),
  heter_fixed  = 0,
  title = "TDT power vs misclassification (pi01)"
)
```


## 6.2 Power sensitivity: heterogeneity


```{r}

tdt_plot_power_heterogeneity(
  N    = 600,
  pd   = 0.30,
  prev = 0.05,
  R1   = 1.5,
  R2   = 2.25,
  alpha = 0.05,
  delta_prime = 1,
  heter_seq      = seq(0, 0.50, by = 0.05),
  misclass_fixed = 0,
  title = "TDT power vs heterogeneity (1 - pi)"
)

```


## 6.3 Required trios: misclassification

```{r}
tdt_plot_sample_size_misclassification(
  target_power = 0.80,
  pd   = 0.30,
  prev = 0.05,
  R1   = 1.5,
  R2   = 2.25,
  alpha = 0.05,
  delta_prime = 1,
  misclass_seq = seq(0, 0.15, by = 0.01),
  heter_fixed  = 0,
  title = "Required trios vs misclassification (pi01)"
)
```

## 6.4 Required trios: heterogeneity

```{r}
tdt_plot_sample_size_heterogeneity(
  target_power = 0.80,
  pd   = 0.30,
  prev = 0.05,
  R1   = 1.5,
  R2   = 2.25,
  alpha = 0.05,
  delta_prime = 1,
  heter_seq      = seq(0, 0.50, by = 0.05),
  misclass_fixed = 0,
  title = "Required trios vs heterogeneity (1 - pi)"
)
```

# 7. 3D Sensitivity Surfaces (optional)

In addition to 2D sensitivity plots, **genmixr** provides optional interactive
3D visualization tools using the `plotly` package.

These functions are only available if `plotly` is installed.

```{r, eval=requireNamespace("plotly", quietly=TRUE)}
tdt_plot3d_power_heterogeneity(
  pd_seq = seq(0.05, 0.80, by = 0.05),
  heter_seq = seq(0, 0.50, by = 0.05),
  N = 600,
  prev = 0.05,
  R1 = 1.5,
  R2 = 2.25
)
```


# 8. Conclusion

This vignette introduced the extended Transmission Disequilibrium Test (TDT) framework implemented in **genmixr**, with emphasis on power calculation, sample size determination, and the consequences of realistic error processes such as phenotype misclassification and locus heterogeneity. 

Through examples based on real studies—Ambrosio et al.’s investigation of DRD2/DRD3 in schizophrenia trios and Spielman et al.’s landmark insulin VNTR TDT paper—we demonstrated how classical trio designs can be re-evaluated and how new studies can be planned under modern assumptions. These examples highlight several key insights:

- **Power can be severely limited** in modest-effect settings, especially when diagnostic uncertainty or genetic heterogeneity is present.  
- Even **small error rates can substantially inflate required sample sizes**, making naive power calculations overly optimistic.  
- The visual tools in **genmixr** (`tdt_plot_power()` and `tdt_plot_sample_size()`) provide intuitive summaries of design robustness, allowing researchers to evaluate sensitivity to multiple forms of error simultaneously.  
- The full modeling framework draws directly from the theoretical development in *Heterogeneity in Statistical Genetics*, which provides a principled foundation for TDT in the presence of mixture structures, LD scaling, and phenotype misclassification.

Together, these tools offer a comprehensive workflow that supports both retrospective interpretation of existing TDT findings and forward-looking design of new trio-based genetic studies. As family-based association methods continue to complement population-based GWAS analyses, incorporating heterogeneity and diagnostic noise into study planning remains essential for reliable inference.

---

# References

Ambrosio AM, Kennedy JL, Macciardi F, Berrettini W, Mutlu C, King N, Muglia P, Vetrugno R, Vincent JB, Macciardi F. **Dopamine D2 and D3 receptor gene polymorphisms in Portuguese schizophrenia trios: No evidence for transmission disequilibrium.** *Psychiatry Research*. 2004;125(3):225–231.  
Available in this vignette’s supporting files.

Spielman RS, McGinnis RE, Ewens WJ. **Transmission test for linkage disequilibrium: The insulin gene region and insulin-dependent diabetes mellitus (IDDM).** *The American Journal of Human Genetics*. 1993;52(3):506–516.  
Available in this vignette’s supporting files.

Gordon D, Finch SJ, Nothnagel M, Ott J. **Heterogeneity in Statistical Genetics.** Oxford University Press; 2020.  
(Primary theoretical reference for the extended TDT framework used in **genmixr**.)

---



